---
title: "Getting started with scholid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with scholid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`scholid` is a lightweight, dependency-free (base R only) toolkit for
working with scholarly and academic identifiers. It provides small,
well-tested helpers to detect, normalize, classify, and extract common
identifier strings.

This vignette introduces the interface and typical workflows for
mixed, messy identifier data.

## Installation

```{r install, eval = FALSE}
install.packages("scholid")
```

## Interface

`scholid` exposes a small set of user-facing functions that operate
consistently across identifier types:

- `scholid_types()` lists supported identifier types.
- `is_scholid(x, type)` checks whether values match the identifier type.
- `normalize_scholid(x, type)` returns canonical identifier strings.
- `extract_scholid(text, type)` extracts identifiers from free text.
- `classify_scholid(x)` guesses the identifier type per element.
- `detect_scholid_type(x)` detects identifier types from canonical or wrapped input values (e.g., URLs or labels).

These generic helpers dispatch internally to type-specific
implementations such as `is_doi()`, `normalize_orcid()`, and
`extract_isbn()`.

## Supported identifier types

```{r scholid types, eval = TRUE}
scholid::scholid_types()
```

## Detect: `is_scholid()`

`is_scholid()` checks whether each value matches a specific identifier
type. It is vectorized and preserves missing values.

```{r check if DOI}
x <- c(
    "10.1000/182",
    "not a doi",
    NA
)
scholid::is_scholid(
    x    = x,
    type = "doi"
)
```

## Normalize: `normalize_scholid()`

Normalization removes common wrappers and enforces a canonical
representation. This is particularly useful when identifiers are stored
as URLs or prefixed labels.

```{r normalize DOI}
x <- c(
  "https://doi.org/10.1000/182.",
  "doi:10.1000/182",
  " 10.1000/182 "
)
scholid::normalize_scholid(
    x    = x, 
    type = "doi"
)
```

For ORCID iDs, normalization removes URL prefixes and enforces hyphenated
grouping.

```{r normalize ORCID}
x <- c(
  "https://orcid.org/0000-0002-1825-0097",
  "0000000218250097"
)
scholid::normalize_scholid(
    x    = x,
    type = "orcid"
)
```

Normalization is designed to be predictable:
- `NA` input stays `NA`.
- Invalid inputs typically become `NA_character_`.

## Extract: `extract_scholid()`

Extraction is for harvesting identifiers from unstructured text. The
result is a list with one element per input element. Each element is a
character vector of matches (possibly empty).

```{r extract DOI}
txt <- c(
  "See https://doi.org/10.1000/182 and doi:10.5555/12345678.",
  "No identifier here.",
  NA
)
scholid::extract_scholid(
    text = txt,
    type = "doi"
)
```

The list return type is intentional: a single text string can contain
multiple identifiers.

## Classify: `classify_scholid()`

`classify_scholid()` returns the best-guess identifier type per element
for mixed identifier columns. Classification is based on the set of
available `is_<type>()` checks and the precedence order defined by
`scholid_types()`.

```{r classify id}
x <- c(
  "10.1000/182",
  "0000-0002-1825-0097",
  "PMC12345",
  "2101.00001v2",
  "not an id",
  NA
)
scholid::classify_scholid(x = x)
```

### Normalization + classification in messy data

Many identifiers appear wrapped (URLs, prefixes, trailing punctuation).
Classification is strict and expects canonical strings. A common
pattern is:

1. Extract identifiers from text.
2. Normalize extracted values.
3. Classify and/or validate.

```{r normalization + classification}
txt <- "Read https://doi.org/10.1000/182 (and ORCID 0000-0002-1825-0097)."
dois <- scholid::extract_scholid(txt, "doi")[[1]]
orcids <- scholid::extract_scholid(txt, "orcid")[[1]]

dois_n <- scholid::normalize_scholid(dois, "doi")
orcids_n <- scholid::normalize_scholid(orcids, "orcid")

scholid::classify_scholid(c(dois_n, orcids_n))
scholid::is_scholid(dois_n, "doi")
scholid::is_scholid(orcids_n, "orcid")
```

## Detect: `detect_scholid_type()`

`detect_scholid_type()` performs best-effort type detection for mixed,
messy identifier input. In contrast to `classify_scholid()`, detection
also recognizes common wrapped forms such as URLs and prefixed labels
(e.g., `doi:`, `https://orcid.org/`, `arXiv:`, `PMID:`).

Detection is useful when working with raw data where identifiers may
not yet be normalized.

For example, wrapped identifiers are not classified strictly:

```{r classification does not work}
x <- c(
  "https://doi.org/10.1000/182",
  "ORCID: 0000-0002-1825-0097",
  "arXiv:2101.00001",
  "PMID: 12345",
  "not an id"
)
scholid::classify_scholid(x)
```

However, they can be detected directly:

```{r detect scholid type}
scholid::detect_scholid_type(x)
```

Whitespace and minor formatting irregularities are handled
conservatively:

```{r detect scholid type with whitespaces}
scholid::detect_scholid_type(
  c(
    " 0000-0002-1825-0097 ",
    " 10.1000/182 ",
    "ISSN 0317-8471"
  )
)
```

`detect_scholid_type()` does not modify values. Once the identifier
type is known, use `normalize_scholid()` to convert to canonical form
and `is_scholid()` for strict validation.

A typical workflow for messy data is:

1. Detect identifier types.
2. Normalize by detected type.
3. Validate canonical identifiers.

This separation keeps detection permissive and normalization
predictable, while preserving strict validation where needed.


## Design notes

`scholid` is intentionally small and conservative:

- It uses base R only at runtime.
- Functions are vectorized and return stable types.
- Type-specific logic is kept in small `is_*()`, `normalize_*()`, and
  `extract_*()` helpers.
- The package is designed to be a low-level building block for other
  packages and for workflows.

## Session information

```{r session info}
sessionInfo()
```
